Type Evolution in Interactive Programming Systems

### 2024-05-30 paper.tex
New kind of diagrams, tweaks

\title{Type Evolution in Programming Systems}

@@

\subsection{Types, Data and Code}

As is the case for buildings, making a change at surface layers (skin or stuff) is relatively easy,
but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
In the case of programs, a more fundamental layer defines the shape of data and code.
We refer to this structure as \emph{types}, although we do not use the term in a precise technical
sense. In our conceptual framework, types may be data types that define the shape of data, they
may be explicitly declared in code and checked statically, as well as dynamically checked or
even implicit. The key property of types for us is that they constrain the shape of other layers
of a program, namely data and code. As a result, a change of program types requires a
corresponding change in program data and code.

The situation is illustrated in Figure~\ref{fig:change} using two variations of a diagram that
we will return to repeatedly in this paper. When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$,
we need to find and apply a corresponding change to other parts of the program whose shape
is defined by the changed type. The problem of \emph{type evolution} is thus finding and
applying changes that will update $\textsf{data}_1$ and $\textsf{code}_1$ into new versions $\textsf{data}_1$ and $\textsf{code}_2$
that correspond to the new type.

The complexity of the problem varies. The first dimension of challenges that we consider
is local. The complexity of the challenge depends on how many other parts of the program
depend on the modified type. If both code and data depend on the type, the complexity of the
program is greater. If there is no associated code or the existing data can be safely discarded,
the complexity is lesser.

The second dimension of challenges that we consider is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. If all variants of the program should
eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
only certain changes from other users (divergence), the complexity is greater. A typical case
may be a combination of the two where types and schema of a program should converge, but data
may diverge.

\begin{figure}[t]
\centering
%\includegraphics[scale=0.175]{figures/diag-data-code.png}
\begin{tikzcd}[sep = 6em]
  \textsf{code}_1
  \arrow[d, "\textsf{refactor}"{name=L, description}]
  &
  \textsf{type}_1
  \arrow[l, "\textsf{constrains}"{above}]
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}"{name=C, description}]
  &
  \textsf{data}_1
  \arrow[d, "\textsf{migrate}"{name=R, description}]
  \\
  \textsf{code}_2
  &
  \textsf{type}_2
  \arrow[l, "\textsf{constrains}"{above}]
  \arrow[r, "\textsf{constrains}"{above}]
  &
  \textsf{data}_2
  \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
\end{tikzcd}
\caption{When a type changes from $\textsf{type}_1$ to $\textsf{type}_2$, a corresponding change needs to take place
  in data and code constrained by the type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

\noindent
Some programming systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when the \texttt{State} type changes. A more effective
programming system based on live programming would allow updates to both the code of the two
functions and the structure of the two types.

@@

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; completed : DateTime option }
type State = { items : Item list }
type Event =
  | .. | SetCompleted of id * DateTime option | ..
\end{lstlisting}

\noindent
To apply the change without restarting the application, the existing data need to be migrated
to match the new type definition. In this case, this likely cannot be done fully automatically
and the programmer may need to provide mapping (\texttt{false} to \texttt{None}, \texttt{true}
to \texttt{Some(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
at the same time, the \texttt{update} function likely does not need to change, but the rendering
code needs to be modified (the programmer needs to specify how to map \texttt{DateTime option}
back to a Boolean for a checkbox).

@@

\frameworkbox{
% \begin{wrapfigure}{r}{14em}
% \vspace{1em}
% \includegraphics[scale=0.175]{figures/diag-both.png}
% \end{wrapfigure}
\frameworkboxtitle{Code-data Co-evolution}
In this case study, the evolving type in Figure~\ref{fig:change} is the \texttt{State} type. When the type
evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
The former is the program logic implemented in \texttt{update} and \texttt{render},
while the latter is the current state (a live value of type \texttt{State}). The scenario
illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
}

\subsection*{Remarks: Requirements and Implementation}
The key requirement from the live programming perspective is to ``do minimal harm''.
The migration of \emph{data} needs to be done automatically and the system should strive
to produce a new state that is as near as possible to the previous state. Migrating
\emph{code} can be done in various ways (transform code, trigger error condition, etc.),
but it is desirable to avoid breaking the \texttt{render} function as this would make the
new application state impossible to visualize.

@@

\section{Entity Evolution in Data-Centric Systems}
Many problems studied in the context of database systems equally apply to programming systems
if we think of them as stateful environments where programs co-exist with their persistent data.
Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

@@

\challenge{Extract Entity}
Acme needs to evolve the above type and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

The system needs to provide a mechanism for referencing entities, such as by using unique
identifiers (here consecutive numbers). This allows, for example, a spelling error in a customer
address to be fixed in one place. It also allow a customer name or address to change without
breaking the connection from all associated orders. Unique identifiers are the essence of
entities, whether they are uniquely generated primary keys in a database or an abstraction
of a memory address in a programming language.

@@

More generally, entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again. Inverse
is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

@@

\frameworkbox{
\begin{wrapfigure}{r}{9em}
\includegraphics[scale=0.2]{figures/diag-variants.png}
\end{wrapfigure}

\frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

\subsection*{Remarks: Representation and Replication}
The central problem is to coordinate the type evolution with code edits to keep the system
executing live and correctly. Solutions may infer the desired type evolution by analysing
source code edits, but would typically offer UI affordances for the commands. They should
try to minimize the number and complexity of commands that must be introduced and also the
interruption to live execution of the code.

@@

\noindent
The same code transformation would need to be applied to queries written over other
data representations such as JSON. The system can also use mechanism such as a database
view and execute the original query over the view. When applying code migrations,
the system should minimize the need for users to manaually intervene. It should also
minimize the possibilities that developer-written code is subtly incorrect in edge cases,
preferably by reducing the need for developer-written code.

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging two independent edits to the data
(Fig.~\ref{fig:conf-add} and Fig.~\ref{fig:conf-tab}); the second merging structural change
with code (Fig.~\ref{fig:conf-budget} and Fig.~\ref{fig:conf-tab}). Formulas in the final version
(Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

@@

\begin{enumerate}
\item Jonathan adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. This changes data, but not the structure (type) of
  the document (Figure~\ref{fig:conf-add}).
\item Tomas refactors the list into a table. He splits the single textual value
  into a name and an email (using a comma as the separator) and adds an additional column for
  tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
  This changes the document structure (type).
\end{enumerate}

\noindent
The system should be able to merge the changes and produce a table containing the new data
in a tabular format (not shown). The result needs to include the additional speaker
created by Jonathan, use the order specified by Jonathan, but use the format defined by
Tomas. The newly added speaker should be reformatted into the new format; for the
``Organizer'' column, the system may use a suitable default value or ask the user interactively.

\frameworkbox{
\begin{wrapfigure}{r}{8em}
\includegraphics[scale=0.25]{figures/diag-data-change.png}
\end{wrapfigure}
\frameworkboxtitle{Independent Data and Code Edits}
In this example, the \emph{type} is implicit and represents the structure of the document.
A change to the \emph{type} done by one organizer needs to be applied to \emph{data} that
has been modified by the other organizer in the meantime, as illustrated on the right.
In the next section, we will see similar pattern, but with formulas, or \emph{code}.
We assume the \emph{decentralized} scenario where \emph{type} evolution can occur in any variant
and \emph{covergence} model where all variants eventually adopt all changes. As discussed
later, the system could also operate in the more challenging \emph{divergence} model.}

@@

Such situations are common in software engineering. Web backend architectures often put data and
code on separate machines, which makes it impossible to update both atomically. To perform
upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscalerails}
to step the database and the code servers forward gradually such that they are compatible with one
another at any given step. One example would be deploying a change to add a database column
before deploying the code that uses the column.

@@

There has been a lot of research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.

>>>>

\title{Type Evolution in Interactive Programming Systems}

@@

\subsection{Types, Data and Code}
In both buildings and programs, making a change at surface layers (skin or stuff) is relatively easy,
but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
In the case of programs, we refer to the more fundamental layer as \emph{types}, although we do not
use the term in a precise technical sense. Types is what defines the shape of data and code.
Types may be database schema, data types that define the shape of data, they may be explicitly
declared in code and checked statically, as well as dynamically checked or even implicit.
The key property of types for us is that they constrain the shape of other layers
of a program, namely data and code. As a result, a change of program types requires a
corresponding change in program data and code.

\begin{wrapfigure}{r}{12em}
\vspace{-1.5em}
\includegraphics[width=12em]{figures/arr-basic.png}
\vspace{-1.5em}
\end{wrapfigure}

The situation is illustrated on the right. We start with code $C_1$ and
data $D_1$ that have a shape corresponding to type $T_1$. When the type changes from $T_1$ to $T_2$ ($\rightarrow$),
a corresponding change needs to take place in data $D_1$ and code $C_1$. The problem
of \emph{type evolution} is synthetising ($\Rightarrow$) suitable corresponding transformations
that will update data $D_1$ and code $C_1$ into new versions $D_2$ and $C_2$ that correspond
to the new type $T_2$.
%
The complexity of the problem varies and finding a suitable transformation may require user
input. If there is code, it typically needs to be transformed, but data can sometimes be
discarded. We refer to such transformations done by a single user as \emph{local} and draw
them inside a box.

\begin{wrapfigure}{r}{12em}
\vspace{-1.5em}
\includegraphics[width=12em]{figures/arr-forkjoin.png}
\vspace{-1.5em}
\end{wrapfigure}

Collaboration adds the \emph{non-local} dimension to the problem. A program may have multiple
variants that exist independently and changes made to them may need to be synchronized. The
situation is illustrated on the right. Here, the initial type $T_1$ is transformed by two
users independently into types $T_2$ and $T_3$. The corresponding data is transformed accordingly,
producing $D_2$ and $D_3$. The challenge is now merging these two diverging transformation and
obtaining a type $T_4$ and data $D_4$ that adopt both of the changes.

The difficulty of the problem depends on what kind of synchronization we want to support.
If all variants of the program should eventually adopt all changes (converge), the complexity
is lesser. If we allow user to apply only certain changes from other users (divergence), the
complexity is greater. A common case may be a combination of the two where types and schema
of a program should converge, but data may diverge.

@@

\noindent
Some systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when the \texttt{State} type changes. A more effective
live programming system would allow live updates to the structure of the two types too.

@@

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; completed : Maybe DateTime }
type State = { items : Item list }
type Event =
  | .. | SetCompleted of id * Maybe DateTime | ..
\end{lstlisting}

\noindent
To apply the change without restarting the application, the existing data need to be migrated
to match the new type definition. In this case, this likely cannot be done fully automatically
and the programmer may need to provide mapping (\texttt{false} to \texttt{Nothing}, \texttt{true}
to \texttt{Just(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
at the same time, the \texttt{update} function likely does not need to change, but the rendering
code needs to be modified (the programmer needs to specify how to map \texttt{DateTime option}
back to a Boolean for a checkbox).

@@

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{0em}
\includegraphics[width=12em]{figures/arr-steps.png}
\vspace{0em}
\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this case study, the evolving type is the \texttt{State} type. When the type evolves,
the program logic (code) and the current live value of the type (data) needs to evolve
correspondingly. The challenge is easier if we see the evolution as a sequence of more primitive
transformations, such as those in Table~\ref{tbl:elmchanges}, that each has a corresponding
primitive code and data transformation. The primitive transformations may either be invoked
through a user interface, or inferred from textual code edits made by the programmer.
}

\subsection*{Remarks: Requirements and Implementation}
The key requirement from the live programming perspective is to ``do minimal harm''.
The migration of \emph{data} needs to be done automatically and the system should strive
to produce a new state that is as near as possible to the previous state. Migrating
\emph{code} can be done in various ways,
but it is desirable to avoid breaking the \texttt{render} function as this would make the
new application state impossible to visualize.

@@

\section{Entity Evolution in Data-Centric Systems}
If we see programming systems as stateful environments where programs co-exist with their persistent
data, we can gain valuable insights from research on database systems.
Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

@@

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

The system needs to provide a mechanism for referencing entities, such as by using unique
identifiers (here consecutive numbers). This allows, for example, a spelling error in a customer
address to be fixed in one place. It also allow a customer name or address to change without
breaking the connection from all associated orders. Unique identifiers are the essence of
entities, whether they are uniquely generated primary keys in a database or an abstraction
of a memory address in a programming language. As spreadsheets do not support relationships
between entities, Acme will need to migrate their data to a database and restructure the original
flat table.

@@

Entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again. Inverse
is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

@@

\subsection*{Remarks: Representation and Replication}
The central problem is to coordinate the schema evolution with code edits to keep the system
executing live and correctly. Solutions may infer the desired schema evolution by analysing
source code edits, but would typically offer UI affordances for the commands. They should
try to minimize the number and complexity of commands that must be introduced and also the
interruption to live execution of the code.

@@

\noindent
In database systems, the system can use a mechanism such as a
view and execute the original query over the view. When applying code migrations,
the system should minimize the need for users to manaually intervene. It should also
minimize the possibilities that developer-written code is subtly incorrect in edge cases,
preferably by reducing the need for developer-written code.

\frameworkbox{
\begin{wrapfigure}{r}{12em}
\vspace{-0.5em}
\includegraphics[width=12em]{figures/arr-db.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Equivalence in Schema Evolution}
We described our conceptual framework in terms of types, code and data, but this case
study shows that the same patterns appear elsewhere. In database systems, the schema
$S_1$ changes into $S_2$ and the data $D_1$ need to transform to corresponding $D_2$.
Our framing shows that queries $Q_1$ also need to transform, which is not always addressed
in work on schema evolution. Extract/Absorb Entity operations are also interesting because
they transform an equivalent pair of schemas and so it should be possible to apply them
(and the corresponding data and query transformations) automatically.
}

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging a data edit
(Fig.~\ref{fig:conf-add}) with a type and data edit (Fig.~\ref{fig:conf-tab}); the second involves
merging a type and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

@@

\begin{enumerate}
\item Jonathan adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. This changes data, but not the structure (type) of
  the document (Figure~\ref{fig:conf-add}).
\item Tomas refactors the list into a table. He splits the single textual value
  into a name and an email (using a comma as the separator) and adds an additional column for
  tracking which of the organizers should contact the speaker (Figure~\ref{fig:conf-tab}).
  This changes the document structure (implicit type).
\end{enumerate}

\noindent
The system should be able to merge the changes and produce a table containing the new data
in a tabular format (not shown). The result needs to include the additional speaker
created by Jonathan, use the order specified by Jonathan, but use the format defined by
Tomas. The newly added speaker should be reformatted into the new format; for the
``Organizer'' column of the newly created table, the system may use a suitable default value
or ask the user interactively.

\frameworkbox{
\begin{wrapfigure}{r}{13em}
\vspace{-0.5em}
\includegraphics[width=13em]{figures/arr-datafork.png}
\vspace{-1em}
\end{wrapfigure}
\frameworkboxtitle{Merging Data and Code Edits}
In this example, the \emph{type} is implicit and represents the structure of the document.
It involves two independent edits. In one (top right), the data $D_1$ is
changed to $D_2'$, but the type $T_1$ remains unchanged. In the other (bottom left), the
type $T_1$ and data $D_1$ co-evolve into $T_2$ and $D_2$ and the data is further changed
to $D_3$. The challenge is to merge the two edits and obtain jointly modified data $D_3'$
of type $T_2$. In the next section, we will see the same pattern, but also involving
code edits. We assume the \emph{convergence} model in this section, but will discuss
the more challenging \emph{divergence} model later in the paper.}

@@

Such situations are common in software engineering. Web backend architectures often put data and
code on separate machines, which makes it impossible to update both atomically. To perform
upgrades, teams must manually perform multi-step zero-downtime deployment processes~\cite{planetscale}
to step the database and the code servers forward gradually such that they are compatible with one
another at any given step. One example would be deploying a change to add a database column
before deploying the code that uses the column.

@@

\textcolor{red}{Maybe insert something about methodology in general?
(ref the table challenges paper? although this paper is no longer so much about challenges)}

There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.
